---
title: "01_QC2"
author: "YasmineIlli"
date: "2026-07-17"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

*(adapted from Lumeng Li's code)*

## Setup

```{r libraries, message = FALSE, warning = FALSE}
library(Seurat)
library(tidyverse)
library(here)
library(clustree)
library(harmony)
library(patchwork)
library(presto)
library(dittoSeq)
```

```{r output_directories, message = FALSE}
# Setup output directories
data_output_directory <- here("output/01_QC/data/QC2/")
plots_output_directory <- here("output/01_QC/plots/QC2/")
RObjects_output_directory <- here("output/01_QC/RObjects/QC2/")
```

```{r plot_settings, message = FALSE}
# Plot settings
theme_set(theme_classic())

# Color palette
pal <- c(
  blue   = "#015EA8",
  rose   = "#B56576",
  blush  = "#FAE8EB",
  navy   = "#0A014F",
  brown  = "#3F220F"
)

# Group colours (4 groups: HC, RaHC, S3, V)
group_cols <- c(V = "#015EA8", RaHC = "#B56576", S = "#3F220F", HC = "#0A014F")
```

```{r message = FALSE}
# Load data
MergedObject <- readRDS(here("output/01_QC/RObjects/QC1/01_QC1_MergedObject.rds"))
```

## Standard pipeline

First check the distribution of nCount_RNA before and after normalization:

```{r normalization, message = FALSE, warning = FALSE}
set.seed(123)

# Reusable histogram function for nCount_RNA distributions
plot_ncount_hist <- function(seurat_obj, plot_title) {
  ggplot(
    seurat_obj[[]],
    aes(x = nCount_RNA)
  ) +
    geom_histogram(
      bins = 20,
      fill = "#B56576",
      color = "#FAE8EB"
    ) +
    theme_minimal() +
    labs(
      title = plot_title,
      x = "nCount_RNA",
      y = "Cell Count"
    )
}

# Check median library size before normalization
paste("Median library size:", median(MergedObject$nCount_RNA))

# Plot nCount_RNA distribution before normalization
hist_pre <- plot_ncount_hist(
  MergedObject,
  "nCount_RNA Distribution (Pre-Normalization)"
)
hist_pre

ggsave(file = file.path(plots_output_directory, "01_QC2_HistogramPreNormalization.png"), 
       plot = hist_pre, width = 6, height = 4, dpi = 300)

# Normalize using median library size as the scale factor
MergedObject <- NormalizeData(
  MergedObject,
  scale.factor = median(MergedObject$nCount_RNA)
)

# Plot nCount_RNA distribution after normalization
# (distribution is unchanged here since scale.factor was already set to the median)
hist_post <- plot_ncount_hist(
  MergedObject,
  "nCount_RNA Distribution (Post-Normalization)"
)
hist_post

# ggsave(file.path(plots_output_directory, "01_QC2_HistogramPostNormalization.png"),
#        plot = hist_post, width = 6, height = 4, dpi = 300)
```

The histogram is exactly the same because NormalizeData() doesn't change the nCount_RNA. We need to use "data" layer instead:

```{r plot normalized, message = FALSE, warning = FALSE}
# Recompute total counts from the normalized layer
norm_counts <- colSums(
  LayerData(MergedObject, layer = "data")
)

# Add to metadata for plotting
MergedObject$nCount_norm <- norm_counts

# Plot normalized totals
hist_norm <- ggplot(
  MergedObject[[]],
  aes(x = nCount_norm)
) +
  geom_histogram(
    bins = 20,
    fill = "#B56576",
    color = "#FAE8EB"
  ) +
  theme_minimal() +
  labs(
    title = "Recomputed Totals from Normalized Layer",
    x = "Sum of log-normalized counts",
    y = "Cell Count"
  )
hist_norm

ggsave(file.path(plots_output_directory, "01_QC2_HistogramPostNormalization.png"), 
       plot = hist_post, width = 6, height = 4, dpi = 300)
```

```{r highly variable genes, message = FALSE, warning = FALSE}
# Highly variable genes
set.seed(123)

# Identify top 2000 variable features using the VST method
MergedObject <- FindVariableFeatures(
  MergedObject,
  selection.method = "vst",
  nfeatures = 2000
)

# Pull top 10 most variable genes
top10 <- head(VariableFeatures(MergedObject), 10)
top10

# Plot variable features and label the top 10
vf_plot <- VariableFeaturePlot(MergedObject)
vf_plot_labeled <- LabelPoints(
  plot = vf_plot,
  points = top10,
  repel = TRUE,
  min.segment.length = 0,     
  segment.color = "#3F220F",   
  segment.size = 0.3,
  box.padding = 0.8,           
  max.overlaps = Inf           
)

vf_plot_labeled

ggsave(filename = file.path(plots_output_directory, "01_QC2_VariableFeatures.png"), 
       plot = vf_plot_labeled, width = 6, height = 4, dpi = 300)
```

```{r scale data, message = FALSE, waring = FALSE}
# Scale Data
set.seed(123)

MergedObject <- ScaleData(MergedObject, 
                          features = rownames(MergedObject))
```

```{r PCA, message = FALSE, warning = FALSE}
# PCA
set.seed(123)

MergedObject <- RunPCA(MergedObject, 
                       features = VariableFeatures(object = MergedObject),
                       approx = FALSE, # to allow reproducibility
                       npcs = 50,
                       verbose = FALSE)

# not important, more to be able to remove low quality cells later
DimPlot(MergedObject, reduction = "pca")
DimHeatmap(MergedObject, dims = 1:18, cells = 500, balanced = TRUE)

# ElbowPlot
elbowPlot <- ElbowPlot(MergedObject, ndims = 50, reduction = "pca")
elbowPlot

ggsave(filename = file.path(plots_output_directory, "01_QC2_ElbowPlot.png"),
       plot = elbowPlot, width = 6, height = 4, dpi = 300)
```

```{r PC selection, message = FALSE, warning = FALSE}
# Automatic PC selection
# Extract PC standard deviations
stdv <- MergedObject[["pca"]]@stdev[1:50]

# Percent variation explained by each PC
percent_stdv <- (stdv / sum(stdv)) * 100
cumulative <- cumsum(percent_stdv)

# Cutoff 1: first PC where cumulative variation exceeds 90% and individual PC explains < 5%
co1 <- which(cumulative > 90 & percent_stdv < 5)[1]

# Cutoff 2: last PC where the drop in variation to the next PC is > 0.05%
co2 <- sort(
  which(diff(percent_stdv) < -0.05),
  decreasing = TRUE
)[1] + 1

co1
co2

# Use the more conservative of the two cutoffs
selected_dim <- min(co1, co2)
paste("selected dimenstions:", selected_dim)
```

```{r UMAP, message = FALSE, warning = FALSE}
# UMAP
set.seed(123)

# Run UMAP on the selected PCs
MergedObject <- RunUMAP(
  MergedObject,
  dims = 1:selected_dim
)

# UMAP before integration: unlabeled vs. coloured by sample
umap_plain <- DimPlot(
  MergedObject,
  reduction = "umap"
)

umap_by_sample <- DimPlot(
  MergedObject,
  reduction = "umap",
  group.by = "orig.ident"
)

umap_pre_integration <- umap_plain | umap_by_sample
umap_pre_integration

ggsave(filename = file.path(plots_output_directory, "01_QC2_umap_pre_integration.png"), 
                            plot = umap_pre_integration, width = 12, height = 5, dpi = 300)
```

## Harmony Integration

```{r Harmony Integration, message = FALSE, warning = FALSE}
set.seed(123)

# Run Harmony
MergedObject <- IntegrateLayers(
  object = MergedObject,
  method = HarmonyIntegration,
  orig.reduction = "pca",
  new.reduction = "harmony",
  group.by.vars = "orig.ident",
  verbose = FALSE
)

# Check if harmony worked
Reductions(MergedObject)
head(MergedObject@meta.data)
```

```{r Plot integration, warning = FALSE, message = FALSE}
# UMAP after Harmony
set.seed(123)

# Run UMAP again
MergedObject <- RunUMAP(MergedObject, 
                        reduction = "harmony",
                        dims = 1:selected_dim,
                        reduction.name = "umap_harmony")

# Post-integration UMAP: unlabeled vs. coloured by sample
umap_harmony_plain <- DimPlot(
  MergedObject,
  reduction = "umap_harmony"
)

umap_harmony_by_sample <- DimPlot(
  MergedObject,
  reduction = "umap_harmony",
  group.by = "orig.ident"
)

umap_post_integration <- umap_harmony_plain | umap_harmony_by_sample
umap_post_integration

ggsave(
  filename = file.path(plots_output_directory, "01_QC2_umap_post_integration.png"),
  plot = umap_post_integration, width = 12, height = 5, dpi = 300
)

```

## Clustering

```{r FindNeighbors and FindClusters, message = FALSE, warning = FALSE}
set.seed(123)

# Combine counts layers of each sample to one
MergedObject[["RNA"]] <- JoinLayers(MergedObject[["RNA"]])

MergedObject <- FindNeighbors(MergedObject,
                              reduction = "harmony",
                              dims = 1:selected_dim)

MergedObject <- FindClusters(MergedObject,
                             resolution = seq(0.1, 0.9, by = 0.1))

head(MergedObject@meta.data)
```

```{r Clustree, message = FALSE, warning = FALSE, fig.height=8, fig.width=12}
set.seed(123)

clustree_plot <- clustree(MergedObject@meta.data[
  , grep("RNA_snn_res",
         colnames(MergedObject@meta.data))
], prefix = "RNA_snn_res.")

clustree_plot

ggsave(filename = file.path(plots_output_directory, "01_QC2_clustree.png"),
       plot = clustree_plot, width = 14, height = 10, dpi = 300)
```

```{r UMAPs at different resolution, message = FALSE, warning = FALSE, fig.height=10, fig.width=12}
set.seed(123)

umap_res_sweep <- DimPlot(
  MergedObject,
  reduction = "umap_harmony",
  group.by = c(
    "RNA_snn_res.0.1",
    "RNA_snn_res.0.2",
    "RNA_snn_res.0.3",
    "RNA_snn_res.0.4",
    "RNA_snn_res.0.5",
    "RNA_snn_res.0.6",
    "RNA_snn_res.0.7",
    "RNA_snn_res.0.8",
    "RNA_snn_res.0.9"
  ),
  ncol = 3
)
umap_res_sweep

ggsave(
  filename = file.path(plots_output_directory, "01_QC2_UMAPdifferentResolutions.png"),
  plot = umap_res_sweep,
  width = 15,
  height = 15,
  dpi = 300
)
```

```{r samples per cluster, message = F, warning = F, fig.height=10, fig.width=12}
set.seed(123)

# All resolution columns to plot
resolutions <- grep(
  "RNA_snn_res",
  colnames(MergedObject@meta.data),
  value = TRUE
)

# Stacked proportion bar of orig.ident within each cluster, one panel per resolution
sample_prop_plots <- lapply(resolutions, function(res) {
  MergedObject@meta.data %>%
    count(
      cluster = .data[[res]],
      orig.ident
    ) %>%
    group_by(cluster) %>%
    mutate(proportion = n / sum(n)) %>%
    ungroup() %>%
    ggplot(
      aes(x = cluster, y = proportion, fill = orig.ident)
    ) +
    geom_col(color = "black", linewidth = 0.2) +
    scale_y_continuous(labels = scales::percent) +
    theme_minimal() +
    labs(title = res, x = NULL, y = "Proportion of cells", fill = "Sample")
})

sample_prop_by_cluster <- wrap_plots(sample_prop_plots, ncol = 2) +
  plot_layout(guides = "collect")
sample_prop_by_cluster

ggsave(
  filename = file.path(plots_output_directory, "01_QC2_SamplePropInClusters.png"),
  plot = sample_prop_by_cluster,
  width = 18,
  height = 16,
  dpi = 300
)
```


## Double proportion plots

We are only going to plot the doublet proportion per cluster to see if there are clusters that should be removed.

```{r doublet proportion, message = FALSE, warning = FALSE}
# Doublet fraction per cluster at the chosen resolution
doublet_by_cluster <- MergedObject@meta.data %>%
  group_by(
    cluster = .data[["RNA_snn_res.0.2"]],
    doublet
  ) %>%
  summarise(
    n = n(),
    .groups = "drop"
  ) %>%
  group_by(cluster) %>%
  mutate(
    proportion = n / sum(n)
  ) %>%
  ungroup()

doublet_by_cluster

write.csv(
  doublet_by_cluster,
  file.path(data_output_directory, "01_QC2_doublet.csv"),
  row.names = FALSE
)

# Stacked bar: doublet vs singlet composition of each cluster
doublet_composition_plot <- ggplot(
  doublet_by_cluster,
  aes(
    x = cluster,
    y = proportion,
    fill = doublet
  )
) +
  geom_col() +
  scale_fill_manual(
    values = c(
      "Singlet" = "#FAE8EB",
      "Doublet" = "#B56576"
    )
  ) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal() +
  labs(
    x = "Cluster (res 0.2)",
    y = "Proportion of cells",
    fill = NULL
  )
doublet_composition_plot

ggsave(
  filename = file.path(plots_output_directory, "01_QC2_doublet_composition_per_cluster.png"),
  plot = doublet_composition_plot,
  width = 8,
  height = 5,
  dpi = 300
)

# Where doublets sit on the UMAP
umap_doublets <- DimPlot(
  MergedObject,
  reduction = "umap_harmony",
  group.by = "doublet",
  order = "Doublet",          # draw doublets on top so they aren't buried
  cols = c(
    "Singlet" = "#FAE8EB",
    "Doublet" = "#B56576"
  )
)
umap_doublets

ggsave(
  filename = file.path(plots_output_directory, "01_QC2_umap_doublets.png"),
  plot = umap_doublets,
  width = 7,
  height = 5,
  dpi = 300
)
```

## Set Resolution

Selected resolution according to expected number of clusters.

```{r set resolution, message = F, warning = F}
MergedObject <- SetIdent(MergedObject,
                         value = MergedObject$RNA_snn_res.0.2)

head(MergedObject@active.ident)
```

```{r umap_selected_resolution, message = F, warning = F, fig.height=8, fig.width=12}
set.seed(123)

umap_selected_res <- DimPlot(MergedObject,
                             reduction = "umap_harmony",
                             group.by = "RNA_snn_res.0.2",
                             label = TRUE)
umap_selected_res

ggsave(filename = file.path(plots_output_directory, "01_QC2_UMAPselectedRes.png"),
       plot = umap_selected_res,
       width = 12, 
       height = 5,
       dpi = 300)

# Visualize per sample
umap_selected_res_PerSample <- DimPlot(MergedObject,
                                      reduction = "umap_harmony",
                                      group.by = "RNA_snn_res.0.2",
                                      split.by = "orig.ident",
                                      label = TRUE)

umap_selected_res_PerSample

ggsave(filename = file.path(plots_output_directory, "01_QC2_UMAPselectedRes_perSample.png"),
       plot = umap_selected_res_PerSample,
       width = 12, 
       height = 5,
       dpi = 300)
```

## Find Markers

```{r find markers, message = F, warning = F}
set.seed(123)

#top 25 marker genes per cluster
MergedObject_Markers <- FindAllMarkers(
  object = MergedObject,
  only.pos = TRUE,
  test.use = "wilcox",
  logfc.threshold = 0.25,
  min.pct = 0.2,
  min.cells.feature = 5,
  return.thresh = 0.05
  )

top10_by_cluster <- MergedObject_Markers %>% 
  dplyr::filter(p_val_adj <= 0.05) %>% 
  dplyr::group_by(cluster) %>% 
  dplyr::slice_max(order_by = avg_log2FC,
                   n = 10,
                   with_ties = TRUE) %>% 
  dplyr::ungroup() %>% 
  dplyr::arrange(cluster, dplyr::desc(avg_log2FC))

#get a preview
print(top10_by_cluster, n = 30)

#make a table
write_csv(top10_by_cluster, 
          file = file.path(data_output_directory, "01_QC2_top10_by_cluster.csv"))
```

```{r marker dotplot, message = FALSE, warning = FALSE, fig.height=9, fig.width=12}
# Top 3 markers per cluster for the dotplot
top3_by_cluster <- MergedObject_Markers %>% 
  filter(p_val_adj <= 0.05) %>% 
  group_by(cluster) %>% 
  top_n(n = 3,
        wt = avg_log2FC)

dotplot_top3 <- dittoDotPlot(MergedObject,
                             vars = unique(top3_by_cluster$gene),
                             group.by = "RNA_snn_res.0.2",
                             size = 5)

ggsave(file.path(plots_output_directory, 
                 "01_QC2_DotPlot_Top3VariableGenes.png"),
       plot = dotplot_top3,
       width = 12,
       height = 9,
       dpi = 300)
```

```{r marker heatmap, message = FALSE, warning = FALSE, fig.height=9, fig.width=12}
top5_by_cluster <- MergedObject_Markers %>% 
  filter(p_val_adj <= 0.05) %>% 
  group_by(cluster) %>% 
  top_n(n = 5,
        wt = avg_log2FC)

heatmap_top5 <- DoHeatmap(MergedObject,
                          features = top5_by_cluster$gene) +
  NoLegend()

heatmap_top5

ggsave(file.path(plots_output_directory, 
                 "01_QC2_Heatmap_Top5VariableFeatures.png"),
       plot = heatmap_top5,
       width = 16,
       height = 12,
       dpi = 300)
```

## Doublet removal

```{r doublet removal, message = FALSE, warning = FALSE}
# Counts before removal
Idents(MergedObject) <- MergedObject@meta.data$doublet
table(MergedObject$doublet)

table(MergedObject@meta.data$orig.ident)

# Save Object before doublet removal
# saveRDS(MergedObject, file.path(RObjects_output_directory,
#                                 "01_QC2_MergedObject_before_doublet_removal.rds"))

# Keep only singlets 
MergedObject <- subset(MergedObject,
                       idents = "Singlet")

table(MergedObject$doublet)

table(MergedObject@meta.data$orig.ident)
```

```{r save object with removed doublets, message = FALSE, warning = FALSE}
saveRDS(
  MergedObject,
  file.path(RObjects_output_directory, "01_QC2_MergedObject_after_doublet_removal.rds")
)
```

```{r}
sessionInfo()
```


