---
title: "01_QC3"
author: "YasmineIlli"
date: "2026-07-22"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

*(adapted from Lumeng Li's code)*

## Setup

```{r libraries, message = FALSE, warning = FALSE}
library(Seurat)
library(tidyverse)
library(here)
library(clustree)
library(harmony)
library(patchwork)
library(presto)
library(dittoSeq)
```

```{r plot settings, message = FALSE, warning = FALSE}
# Colour palette
palette_primary   <- "#015EA8"   # presentation text only, use minimally
palette_mauve     <- "#B56576"   # main fill colour
palette_blush     <- "#FAE8EB"   # secondary / light fill
palette_navy      <- "#0A014F"   # contrast fill
palette_brown     <- "#3F220F"   # contrast fill

# Discrete palette for categorical plots (clusters, samples, groups)
palette_discrete <- c(palette_mauve,
                      palette_navy,
                      palette_brown,
                      palette_blush,
                      palette_primary)

# Continuous ramp: light blush -> mauve -> navy
palette_continuous <- c(palette_blush,
                        palette_mauve,
                        palette_navy)

theme_set(theme_minimal())
```

```{r output directories, message = FALSE, warning = FALSE}
data_output_directory <- here("output/01_QC/data/QC3/")
plots_output_directory <- here("output/01_QC/plots/QC3/")
RObjects_output_directory <- here("output/01_QC/RObjects/QC3/")
```

```{r load data}
MergedObject <- readRDS(here("output/01_QC/RObjects/QC2/01_QC2_MergedObject_after_doublet_removal.rds"))
```

```{r clean meta data}
#remove unused resolutions, clean up meta.data
MergedObject@meta.data$RNA_snn_res.0.1 <- NULL
MergedObject@meta.data$RNA_snn_res.0.2 <- NULL
MergedObject@meta.data$RNA_snn_res.0.4 <- NULL
MergedObject@meta.data$RNA_snn_res.0.5 <- NULL
MergedObject@meta.data$RNA_snn_res.0.6 <- NULL
MergedObject@meta.data$RNA_snn_res.0.7 <- NULL
MergedObject@meta.data$RNA_snn_res.0.8 <- NULL
MergedObject@meta.data$RNA_snn_res.0.9 <- NULL
MergedObject@meta.data$RNA_snn_res.1 <- NULL

head(MergedObject@meta.data)
```

## Check data

```{r median library size}
# Median library size - used below as the normalisation scale factor
median_counts <- median(MergedObject$nCount_RNA)
median_counts
```

```{r plotting normalization}
p <- ggplot(MergedObject@meta.data,
            aes(x = nCount_RNA)) +
  geom_histogram(bins = 20,
                 fill = palette_mauve,
                 colour = "white") +
  labs(x = "nCount_RNA",
       y = "Count")

p

ggsave(filename = file.path(plots_output_directory, "01_QC3_nCount_histogram.png"),
       plot = p,
       width = 8,
       height = 6)
```

We are going to normalize the data again because doublet removal and integration can change the data. But first we need to check if what we would be normalizing. We only want to normalize raw counts not already normalized data, since this normalizing normalized data would distort the values.

```{r check normalization}
range(LayerData(MergedObject,
                layer = "counts")@x)
```

The output should be whole numbers e.g. 0 4523. In that case we can normalize the data. If it's not whole numbers e.g. 0.00 12.743 the values are already normalized. Normally NormalizeData() takes counts and puts normalized data in data slots, so it should be fine, but sometimes it can be corrupted so it is always safer to check.

## Standard Pipeline

```{r normalization, message = FALSE}
MergedObject <- NormalizeData(MergedObject,
                              scale.factor = median(MergedObject$nCount_RNA))
```

Now we can visualize the counts and data before and after normalization. Normalization does not change nCount_RNA so doing a histogram of nCount_RNA (like Lumeng did) is not useful for visualizing normalization.

```{r histogram bf and af normalization}
# Subsample non-zero entries from both layers for a fast distribution comparison
n_sample <- 1e5

raw_values <- LayerData(MergedObject,
                        layer = "counts")@x
norm_values <- LayerData(MergedObject,
                         layer = "data")@x

expression_df <- bind_rows(tibble(expression = sample(raw_values,
                                                      n_sample),
                                  type = "Raw counts"),
                           tibble(expression = sample(norm_values,
                                                      n_sample),
                                  type = "Normalized data")) |>
  mutate(type = fct_relevel(type,
                            "Raw counts",
                            "Normalized data"))

p <- ggplot(expression_df,
            aes(x = expression)) +
  geom_histogram(bins = 50,
                 fill = palette_mauve,
                 colour = "white") +
  facet_wrap(~ type,
             scales = "free") +
  labs(x = "Expression",
       y = "Count")

p

ggsave(filename = file.path(plots_output_directory, "01_QC3_nCount_histogram_bf_af_normalization.png"),
       plot = p,
       width = 10,
       height = 5)
```

Raw counts usually have an extreme right skew with most values near 0 and a long tail. Normalized data should be compressed and more symmetric with values roughly from 0-10.

Now we are going to run all standard steps again, because each step that changes the number of cells (like doublet removal) can have downstream effects, so we should do it again to get the final qc-ed object.

```{r find variable features, message = FALSE}
MergedObject <- FindVariableFeatures(MergedObject,
                                     selection.method = "vst",
                                     nfeatures = 2000)
top10 <- head(VariableFeatures(MergedObject), 10)
top10
```

```{r variable feature plot}
vf_plot <- VariableFeaturePlot(MergedObject)
vf_plot <- LabelPoints(plot = vf_plot,
            points = top10,
            repel = TRUE)
vf_plot

ggsave(filename = file.path(plots_output_directory, "01_QC3_VariableFeatures.png"),
       plot = vf_plot,
       width = 8,
       height = 6)
```

```{r scaling, message = FALSE, warning = FALSE}
set.seed(123)
MergedObject <- ScaleData(MergedObject,
                          features = VariableFeatures(MergedObject), # Changed this from all genes to only variablefeatures to make it computationally less expensive (run faster and less memory used) and the only downstream process using it for now is PCA which will run on variablefeatures only anyways
                          vars.to.regress = c("percent.mt",
                                              "percent.hb"))
```

```{r PCA, message = FALSE}
set.seed(123)
#approx=FALSE for reproducibility regardless of the CPU
#because approx=FALSE calculates all different PCs we set a max to npcs=50, because we do not need more than 1000
MergedObject <- RunPCA(MergedObject,
                       features = VariableFeatures(object = MergedObject),
                       verbose = FALSE,
                       approx = FALSE,
                       npcs = 50,
                       reduction.name = "pca_DR")

p <- DimPlot(MergedObject,
             reduction = "pca_DR")

p

ggsave(filename = file.path(plots_output_directory, 
                            "01_QC3_PC_DimPlot.png"),
       plot = p,
       width = 10)
```

```{r PC heatmap, message = FALSE}
set.seed(123)
p <- DimHeatmap(MergedObject,
                 reduction = "pca_DR",
                 dims = 1:18,
                 cells = 500,
                 balanced = TRUE,
                 fast = FALSE # To make it return a ggplot object
                ) &
  scale_fill_gradient2(low = palette_navy,
                       mid = palette_blush,
                       high = palette_mauve,
                       midpoint = 0)

p

ggsave(filename = file.path(plots_output_directory, 
                            "01_QC3_PC_Heatmap.png"),
       plot = p,
       height = 20,
       width = 18,
       dpi = 900)
```

```{r ElbowPlot, message = FALSE}
p <- ElbowPlot(MergedObject,
               ndims = 50,
               reduction = "pca_DR")

p

ggsave(filename = file.path(plots_output_directory, 
                            "01_QC3_ElbowPlot.png"),
       plot = p)
```

PC selection criteria:

-   co1: number of PCs reaching >90% cumulative variance while each individual PC contributes <5%

-   co2: the last PC that still shows a >0.05% drop in variance relative to the previous PC

The lower of the two is used.

```{r PC selection}
# approx = FALSE returns stdev for all PCs regardless of npcs, so slice to 50
stdv <- MergedObject[["pca_DR"]]@stdev[1:50]
percent_stdv <- (stdv/sum(stdv)) * 100

cumulative <- cumsum(percent_stdv)

# co1: cumulative variance above 90% and individual contribution below 5%
co1 <- which(cumulative > 90 & percent_stdv < 5)[1]

# co2: last PC with a drop in variance greater than 0.05 percentage points
co2 <- sort(which((percent_stdv[1:length(percent_stdv) - 1] - 
                     percent_stdv[2:length(percent_stdv)]) >
                    0.05),
            decreasing = T)[1] + 1

#select the number of PCs (dims)
selected_dim <- min(co1, co2)

c(co1 = co1,
  co2 = co2,
  selected_dim = selected_dim)
```

## Harmony integration

```{r Harmony, message = FALSE}
set.seed(123)

MergedObject <- RunHarmony(object = MergedObject,
                           group.by.vars = "orig.ident",
                           reduction = "pca_DR",
                           reduction.save = "harmony_DR",
                           verbose = FALSE,
                           dims.use = 1:selected_dim)

#Check that harmony reduction is saved and the seurat object is ok
Reductions(MergedObject)
```

```{r run UMAP, message = FALSE, warning = FALSE}
set.seed(123)

# reduction.name keeps the pre-doublet-removal UMAPs intact for comparison below
MergedObject <- RunUMAP(MergedObject,
                        reduction = "harmony_DR",
                        dims = 1:selected_dim,
                        reduction.name = "umap_harmony_DR")
```

We can compare the umaps on the doublet removed data set with the RunUMAP before integration and doublet removal, with RunUMAP after integration and before doublet removal and the RunUMAP after integration and doublet removal:

```{r UMAP comparison, message = FALSE}
# Build all six panels in one pass over the reduction names
umap_reductions <- c("umap",
                     "umap_harmony",
                     "umap_harmony_DR")

umap_panels <- map(umap_reductions,
                   ~ DimPlot(MergedObject,
                             reduction = .x,
                             cols = palette_discrete) |
                     DimPlot(MergedObject,
                             reduction = .x,
                             group.by = "orig.ident"))

p <- wrap_plots(umap_panels,
                ncol = 1)

p

ggsave(filename = file.path(plots_output_directory, 
                            "01_QC3_UMAP_comparison.png"),
       plot = p,
       width = 14,
       height = 18)
```

## Clustering

```{r Clustering, message = FALSE}
set.seed(123)

MergedObject <- FindNeighbors(MergedObject,
                              reduction = "harmony_DR",
                              dims = 1:selected_dim)

MergedObject <- FindClusters(MergedObject,
                             resolution = seq(0.1, 
                                              1.0,
                                              by = 0.1))
head(MergedObject@meta.data)
```

```{r Clustree, message = FALSE}
p <- clustree(MergedObject@meta.data[, grep("RNA_snn_res",
                                       colnames(MergedObject@meta.data))],
              prefix = "RNA_snn_res.")
p

ggsave(filename = file.path(plots_output_directory,
                            "01_QC3_clustree.png"),
       plot = p)
```

```{r DimPlot per resolution, message = FALSE, fig.width = 16, fig.height = 10}
set.seed(123)
p <- DimPlot(MergedObject,
             reduction = "umap_harmony_DR",
             group.by = c("RNA_snn_res.0.1",
                          "RNA_snn_res.0.2",
                          "RNA_snn_res.0.3", 
                          "RNA_snn_res.0.4", 
                          "RNA_snn_res.0.5", 
                          "RNA_snn_res.0.6", 
                          "RNA_snn_res.0.7", 
                          "RNA_snn_res.0.8", 
                          "RNA_snn_res.0.9",
                          "RNA_snn_res.1"))
p & NoLegend()

ggsave(filename = file.path(plots_output_directory, 
                            "01_QC3_DimPlotPerRes.png"),
       height = 10,
       width = 16)
```

```{r DimPlot selected resolution, message = FALSE, fig.width = 14, fig.height= 8}
selected_resolution <- "RNA_snn_res.0.3"

p <- DimPlot(MergedObject,
             reduction = "umap_harmony_DR",
             group.by = selected_resolution,
             label = TRUE) |
      DimPlot(MergedObject,
              reduction = "umap_harmony_DR",
              group.by = "orig.ident")
p

ggsave(filename = file.path(plots_output_directory, 
                            "01_QC3_DimPlotSelectedRes.png"),
       width = 14,
       height = 8)
```

## Adding patient groups

We are going to add the individual patients to the following groups:

-   HC: healthy controls

-   Pre: preSSc

-   S: eSSc

-   PRP: Primary Raynauds

```{r add patient groups}
MergedObject@meta.data <- MergedObject@meta.data |>
  mutate(group.ident = case_when(grepl("^RaHC", orig.ident) ~ "PRP",
                                 grepl("^HC",   orig.ident) ~ "HC",
                                 grepl("^V",    orig.ident) ~ "Pre",
                                 grepl("^S",    orig.ident) ~ "S",
                                 TRUE ~ NA_character_))

# Verify every cell was assigned and the mapping is as expected
table(MergedObject$orig.ident,
      MergedObject$group.ident)
```

## UMAPs and QC Plots

```{r UMAP split by group, message=F, fig.height=8, fig.width=14}
p <- DimPlot(MergedObject,
             reduction = "umap_harmony_DR",
             group.by = selected_resolution,
             split.by = "group.ident")
p

ggsave(filename = file.path(plots_output_directory, 
                            "01_QC3_DimPlotperGroup.png"),
       plot = p,
       width = 14,
       height = 8)
```

```{r UMAP groups, message=F, fig.height=8, fig.width=14}
#Check if the group idents are correct
p <- DimPlot(MergedObject,
             reduction = "umap_harmony_DR",
             group.by = "group.ident",
             split.by = "group.ident",
             cols = palette_discrete)
p

ggsave(filename = file.path(plots_output_directory, 
                            "01_QC3_DimPlotGroups.png"),
       plot = p,
       width = 14,
       height = 8)
```

```{r final QC metrics, message=F, fig.width=12, fig.height=8}
p <- VlnPlot(MergedObject,
             group.by = selected_resolution,
             pt.size = 0,
             features = c("nFeature_RNA",
                          "nCount_RNA",
                          "percent.mt",
                          "percent.hb",
                          "percent.ribo"),
             ncol = 2)
p

ggsave(filename = file.path(plots_output_directory, 
                            "01_QC3_VlnPlotFinalQCMetrics.png"),
       plot = p,
       width = 12,
       height = 8)
```

```{r QC metrics per sample, message=F, fig.width=12, fig.height=8}
p <- VlnPlot(MergedObject,
             group.by = "orig.ident",
             pt.size = 0,
             features = c("nFeature_RNA",
                          "nCount_RNA",
                          "percent.mt",
                          "percent.hb",
                          "percent.ribo"),
             ncol = 2)
p

ggsave(filename = file.path(plots_output_directory, 
                            "01_QC3_VlnPlotFinalQCMetrics_perSample.png"),
       plot = p,
       width = 12,
       height = 8)
```

```{r QC metrics per group, message=F, fig.width=12, fig.height=8}
p <- VlnPlot(MergedObject,
             group.by = "group.ident",
             pt.size = 0,
             features = c("nFeature_RNA",
                          "nCount_RNA",
                          "percent.mt",
                          "percent.hb",
                          "percent.ribo"),
             ncol = 2)
p

ggsave(filename = file.path(plots_output_directory, 
                            "01_QC3_VlnPlotFinalQCMetrics_perGroup.png"),
       plot = p,
       width = 12,
       height = 8)
```

## Summary

Cells per sample:

```{r cells per sample}
# Cell counts per sample and per group in one table
cell_counts <- MergedObject@meta.data %>% 
  count(orig.ident,
        group.ident,
        name = "n_cells") %>% 
  arrange(group.ident,
          orig.ident)

cell_counts
```

Cells per group:

```{r cells per group}
cell_counts_perGroup <- cell_counts %>% 
  group_by(group.ident) %>% 
  summarize(sum(n_cells)) %>% 
  ungroup() %>% 
  mutate(
    sample_size = case_when(
      group.ident == "HC" ~ 6,
      group.ident == "PRP" ~ 4,
      group.ident == "Pre" ~ 9,
      group.ident == "S" ~ 7
    )
  )

cell_counts_perGroup
```

```{r total cells}
total_cells <- sum(cell_counts$n_cells)
```

Total number of cells: `r total_cells`.

```{r save object}
saveRDS(MergedObject,
        file = file.path(RObjects_output_directory,
                         "01_QC3_FinalMergedObject.rds"))
```

## SessionInfo

```{r}
sessionInfo()
```

